Mixed-Effects Model for MTLD by Gender

### 2018-08-23 paper.Rmd
merged all previous changes from merge issue in Git; small grammar fixes; clarified role of CLANtoR; clarified description of CHAT tiers

author_note: |
  Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work. This work is supported by a Jacobs Advanced Research Fellowship to MCF.

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development, particularly in characterizing the early language learning environment.
  Access to these data can be both complex for novices and difficult to automate for advanced users, however.
  To address these issues, we introduce `childes-db`, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
  Along with versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.

@@

Since its inception, a tremendous body of research has made use of CHILDES data.
Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008], and many others.
Some studies analyze individual examples in depth [e.g., @snyder2007], others track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of all child or caregiver speech pooled across corpora [e.g., @redington1998; @montagEtAl2015].
<!-- These are just a few of the thousands of studies that make use of CHILDES corpora. -->

Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers stored on separate lines.
These tiers allow information about utterances to be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance.
These files are usually analyzed using a command-line program (CLAN) that allows users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool with its many functions and flags can be daunting.
These technical barriers decrease the ease of exploration by a novice researcher or in a classroom exercise.

On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in Python or R.
Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].

The practice of writing custom parsers is problematic for a number of reasons.
First, effort is wasted in implementing the same features again and again.
Second, this process can introduce errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
Third, these parsing scripts are rarely shared -- and when when they are, they typically break with subsequent revisions to the dataset -- leading to much greater difficulty in reproducing the exact numerical results from previous published research that used CHILDES [see e.g., @meylan2017 for an example].
Fourth, the CHILDES corpus itself is a moving target: computational work using the entire corpus at one time point may include a different set of data than subsequent work due as corpora are added and revised. 
Currently, there is no simple way for researchers to document exactly which version of the corpus has been used, short of creating a full mirror of the data.
These factors together lead to a lack of *computational reproducibility*, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].

@@

We begin by describing the architecture of `childes-db` and the web applications that we provide.
Next, we describe the `childesr` API, which provides a set of R functions for programmatic access to the data while abstracting away many of the technical details.
We conclude by presenting several worked examples of specific uses of the system -- both web apps and the R API -- for research and teaching.

@@

As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users, often using CLAN.
In contrast to this parsing approach, which entails the sequential processing of strings, `childes-db` treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts [see @klineCLANtoR for an earlier example of deriving tabular representations of CHILDES].
Users of data analysis languages like R or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tables such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
These operations obviate the need for users to consider the specifics of the CHAT representation -- instead they simply request the entities they need for their research and allow the API to take care of the formatting details.
We begin by orienting readers to the design of the system via a top-level description and motivation for the design of the database schema, then provide details on the database’s current technical implementation and the versioning scheme.
Users primarily interested in accessing the database can skip these details and focus on access through the `childesr` API and the web apps.
<!-- SM: not sure if we want to have readers skip this; there may be some conceptual dependencies here -->
<!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->

@@

At its core, `childes-db` is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
The smallest unit of abstraction tracked by the database is a *token*, treated here as the standard (or citation) orthographic form of a word.
Using the standardized written form of the word facilitates the computation of lexical frequency statistics for comparison or aggregation across children or time periods.
Deviations from the citation form -- which are particularly common in the course of language development and often of interest to researchers -- are kept as a separate (possibly null) field associated with each token.

@@

Many of the other tables in the database are hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- that store attributes appropriate for each level of description.
Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
An *utterance* contains one or more words and includes fields such as the utterance type such as *declarative* or *interrogative*, total number of tokens, and the total number of morphemes if the morphological structure is available in the original CHAT file.
A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, and age in days if defined, and the filename from CHILDES.
A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown [@brown1973] or Providence [@demuth2006word] corpora.
Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
Because every record can be linked to a top-level collection (generally corresponding to a language), each table includes data from all languages represented in CHILDES.

Participants -- generally children and caregivers -- are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages (or age of corresponding child) for which they are observed.
For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data corresponding to children and their caregivers can be easily associated.

@@

`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database server that can be accessed directly from a wide range of programming languages.
The `childes-db` project provides hosted, read-only databases for direct access and for `childesr` (described below) as well as compressed .sql exports for local installation.
While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.

@@

The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked.
To facilitate reproducibility of past analyses, we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
By default, users interact with the most recent version of the database available.
To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server.
<!-- SM: How easy is it to implement versioning in the Shiny apps? This would show anyone using the web app the availability of the versions AS: app functionality depends on db version in some cases, so nontrivial SM: let's punt-->

@@

We will prioritize the addition of these information sources and others in response to community feedback. <!--AS: removed this, as adding this info would require adding new columns: extending coverage to the above information sources will not require changes to the existing database schema. -->

@@

We first discuss the `childes-db` web apps and then introduce the `childesr` R package.

@@

The ability to easily browse and explore the CHILDES corpora is a cornerstone of the `childes-db` project.
To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
All of this is doable without the requirement of understanding command-line tools or any kind of programming knowledge as had been the case with CLAN.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but making the outputs into browsable visualizations.]

Our web apps are built using Shiny, an R package that enables easy app construction using R.
Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular R plotting package `ggplot2`.
A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot.
All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easily extract developmental indices quickly and without any technical overhead.
<!-- We next walk through the three web apps provided on our website and describe their functionality in more detail.  -->

@@

Several of the most common of these measures are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
As with the Frequency Counts app, caregivers' lexical diversity measures can be plotted alongside children's.

We have currently implemented the following measures:

@@

Many times a researcher will want to investigate the statistics of corpora (e.g., their size, number of utterances, number of tokens) before choosing a target corpus or set of corpora for a project.
This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as giving researchers insight into the aggregate characteristics of CHILDES.
For example, examining the aggregate statistics reveals that coverage in CHILDES peaks at around 30 months.
<!-- SM: note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->

@@

Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
Making use of the R programming language [@R-base], we provide the `childesr` package. R is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g. @song2015; @norrman2015].
The `childesr` package abstracts away the details of connecting to and querying  the database.
Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which makes manipulating large datasets quick and easy.
We describe the commands that the package provides and then give several worked examples of using the package for analyses.

The `childesr` package is easily installed via CRAN, the comprehensive R archive network.
To install, simply type: `install.packages("childesr")`. After installation, users have access to functions that can be used to retrieve tabular data from the database:

@@

In this section we give a number of examples of how `childes-db` can be used in both research and teaching, using both the web apps and the R API. Note that all of these examples use `dplyr` syntax [@R-dplyr]; several accessible introductions to this framework are available online [e.g., @wickham2016].

@@

```{r color-shiny, out.width = "500px", fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development"}
knitr::include_graphics("images/color-shiny.png")
```

@@

For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
We store these in the `colors` variable, and then use the `get_types()` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the North American English corpora.

@@

We now join these two pieces of information together -- how many times each speaker produced each color word, and how many total words they produced.
We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
For comparability with the web app analysis, these proportions are converted to parts per million words.

@@

```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age. Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap.", cache = TRUE}
ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat,
                            ymin = ci_lower, ymax = ci_upper,
                            color = color, label = color)) +
  geom_pointrange(position = position_dodge(.5)) +
  geom_point() +
  geom_smooth(method = "loess", se = F, span = .5) +
  facet_wrap(~ color, scales = "free_y") +
  theme_few() +
  scale_color_manual(values = graph_colors$graph_color) +
  scale_x_continuous(breaks = seq(12, 60, 12),
                     name = "Age (months)") +
  scale_y_continuous(name = "Frequency (per million words)") +
  geom_dl(method = list(dl.trans(x = x + .2), "last.qp", cex = 1)) +
  theme(legend.position = "none")
```

@@

```{r echo=TRUE}
mtld_data <- speaker_stats %>%
  filter(!is.na(target_child_sex), target_child_name != "Leo",
         language %in% included_languages) %>%
  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
  summarise(measure = mean(mtld)) %>%
  ungroup() %>%
  mutate(age_years = target_child_age / 12,
         target_child_sex = factor(target_child_sex,
                                   levels = c("male", "female"))) %>%
  filter(age_years < max_age)
```

@@

```{r}
mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
              (1 | target_child_id) +
              (target_child_sex + age_years | language),
            data = mtld_data, REML = TRUE, control = lmerControl(
              optimizer = "optimx", optCtrl = list(method = "nlminb")))

@@

mod_data$measure <- predict(
  mod, newdata = mod_data, re.form = ~(target_child_sex + age_years | language)
)
```

@@

## Teaching with `childes-db`

@@

1. introduce the research question (e.g., What are the types of words that children first produce?),
2. allow students to reflect or do a pair-and-share discussion with their neighbor,
3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel a of Figure \ref{fig:in-class-demo-plot}),
4. elicit hypotheses from students about the kinds of words that children are likely to produce,
5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels b and c of Figure \ref{fig:in-class-demo-plot}), and
6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel d of Figure \ref{fig:in-class-demo-plot}).

```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching. Panels a-d show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}

@@

We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
This database -- together with the R API and web apps -- facilitates the use of child language data.
For teachers, students, and casual explorers, the web apps allow browsing and demonstration.
For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data.
We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.

>>>>

author_note: |
   Co-first authorship indicated with $*$. Thanks to Brian MacWhinney for advice and guidance, and to Melissa Kline for her work on ClanToR, which formed a starting point for our work. This work is supported by a Jacobs Advanced Research Fellowship to MCF.

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development, particularly in characterizing the early language learning environment.
  Access to these data can be both complex for novices and difficult to automate for advanced users, however.
  To address these issues, we introduce `childes-db`, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an `R` application programming interface (API).
  Along with versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.

@@

Since its inception, a tremendous body of research has made use of CHILDES data.
Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008], and many others.
Some studies analyze individual examples in depth [e.g., @snyder2007], some track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of all child or caregiver speech pooled across corpora [e.g., @redington1998; @montagEtAl2015].
<!-- These are just a few of the thousands of studies that make use of CHILDES corpora. -->

Nonetheless, there are some outstanding challenges working with CHILDES, both for students and for advanced users.
The CHILDES ecosystem uses a specialized file format (CHAT), which is stored as plain text but includes structured annotations grouped into parallel information ``tiers'' on separate lines.
These tiers allow for a searchable plaintext transcript of an utterance to be stored along with structured annotations of its phonological, morphological, or syntactic content.
These files are usually analyzed using a command-line program (CLAN) that allows users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execute complex searches against the data.
While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool with its many functions and flags can be daunting.
These technical barriers decrease the ease of exploration by a novice researcher or in a classroom exercise.

On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in Python or R.
Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analysis typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @klineCLANtoR; @redington1998;@yang2013;@meylan2017].

The practice of writing custom parsers is problematic for a number of reasons.
First, effort is wasted in implementing the same features again and again.
Second, this process can introduce errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
Third, these parsing scripts are rarely shared -- and when when they are, they typically break with subsequent revisions to the dataset -- leading to much greater difficulty in reproducing the exact numerical results from previous published research that used CHILDES [see e.g., @meylan2017 for an example].
Fourth, the CHILDES corpus itself is a moving target: computational work using the entire corpus at one time point may include a different set of data than subsequent work as corpora are added and revised. 
Currently, there is no simple way for researchers to document exactly which version of the corpus has been used, short of creating a full mirror of the data.
These factors together lead to a lack of *computational reproducibility*, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].

@@

We begin by describing the architecture of `childes-db` and the web applications that we provide.
Next, we describe the `childesr` API, which provides a set of `R` functions for programmatic access to the data while abstracting away many of the technical details.
We conclude by presenting several worked examples of specific uses of the system -- both web apps and the `R` API -- for research and teaching.

@@

As described above, CHILDES is most often approached as a set of distinct CHAT files, which are then parsed by users, often using CLAN.
In contrast to this parsing approach, which entails the sequential processing of strings, `childes-db` treats CHILDES as a set of linked tables, with records corresponding to intuitive abstractions such as words, utterances, and transcripts [see @klineCLANtoR for an earlier example of deriving a singular tabular representation of a CHILDES transcript].
Users of data analysis languages like `R` or Julia, libraries like Pandas, or those familiar with Structured Query Language (SQL) will be familiar with operations on tabular representations of data such as filtering (subsetting), sorting, aggregation (grouping), and joins (merges).
These operations obviate the need for users to consider the specifics of the CHAT representation -- instead they simply request the entities they need for their research and allow the API to take care of the formatting details.
We begin by orienting readers to the design of the system via a top-level description and motivation for the design of the database schema, then provide details on the database’s current technical implementation and the versioning scheme.
Users primarily interested in accessing the database can skip these details and focus on access through the `childesr` API and the web apps.
<!-- SM: not sure if we want to have readers skip this; there may be some conceptual dependencies here -->
<!-- We first present the design and technical approach of childes-db, then describe the interactive analysis tools provided by our web-based frontend.  -->

@@

At its core, `childes-db` is a database consisting of a set of linked tabular data stores where records correspond to linguistic entities like words, utterances, and sampling units like transcriptions and corpora.
The smallest unit of abstraction tracked by the database is a *token*, treated here as the standard (or citation) orthographic form of a word.
Using the standardized written form of the word facilitates the computation of lexical frequency statistics for comparison or aggregation across children or time periods.
Deviations from the citation form -- which are particularly common in the course of language development and often of special interest to researchers -- are kept as a separate (possibly null) field associated with each token.

@@

Many of the other tables in the database describe hierarchical collections built out of tokens -- *utterance*, *transcript*, *corpus*, and *collection* -- and store attributes appropriate for each level of description.
Every entity includes attributes that link it to all higher-order collections, e.g., an utterance lists the transcript, corpus, and collection to which it belongs.
An *utterance* contains one or more word tokens and includes fields such as the utterance type (e.g.,  *declarative*, *interrogative*, etc.), total number of tokens, and the total number of morphemes if the morphological structure is available in the original CHAT file.
A *transcript* consists of one or more utterances and includes the date collected, the name of the target child, the age in days if defined, and the filename from CHILDES.
A *corpus* consists of one or more transcripts, corresponding to well-known collections like the Brown [@brown1973] or Providence [@demuth2006word] corpus.
Finally, a *collection* is a superordinate collection of corpora generally corresponding to a geographic region, following the convention in CHILDES.
Because every record can be linked to a top-level collection (generally corresponding to a language), each table includes data from all languages represented in CHILDES.

Participants -- generally children and caregivers -- are represented separately from the token hierarchy because it is common for the same children to appear in multiple transcripts.
A participant identifier is associated with every word and utterance, including a name, role, 3-letter CHILDES identifier (CHI = child, MOT = mother, FAT = father, etc.), and the range of ages for which they are observed (or age of corresponding child, in the case of caregivers).
For non-child participants (caregivers and others), the record additionally contains an identifier for the corresponding target child, such that data corresponding to children and their caregivers can be easily associated.

@@

`childes-db` is stored as a MySQL database, an industry-standard, open-source relational database that can be accessed directly from a wide range of programming languages.
The `childes-db` project provides access to hosted, read-only databases on a publicly-accessible server for direct access and `childesr` (described below).
The project also hosts compressed .sql exports for local installation.
While the former is appropriate for most users, local installation can provide performance gains by allowing a user to access the database on their machine or on their local network, as well as allowing users to store derived information in the same database.

@@

The content of CHILDES changes as additional corpora are added or transcriptions are updated; as of time of writing, these changes are not systematically tracked in a public repository.\footnote{Specific versions of the database, tracked using the version control system Git, can be obtained by emailing the maintainers of the CHILDES project.
While tracking line-level changes with Git provides detailed information about what has changed, our method allows researchers to access the relevant version programatically by simply adding an argument to a function call.}
To facilitate reproducibility of past analyses, we introduce a simple versioning system by adding a new complete parse of the current state of CHILDES every six months or as warranted by changes in CHILDES.
By default, users interact with the most recent version of the database available.
To support reproduction of results with previous versions of the database, we continue to host recent versions (up to the last three years / six versions) through our `childesr` API so that researchers can run analyses against specific historical versions of the database.
For versions more than three years old, we host compressed .sql files that users may download and serve using a local installation of MySQL server (for which we provide instructions).
<!-- SM: How easy is it to implement versioning in the Shiny apps? This would show anyone using the web app the availability of the versions AS: app functionality depends on db version in some cases, so nontrivial SM: let's punt-->

@@

\noindent At present, `childes-db` focuses strictly on the contents of CHILDES, and does not include material in related TalkBank projects such as PhonBank, AphasiaBank, or DementiaBank.
We will prioritize the addition of these information sources and others in response to community feedback. <!--AS: removed this, as adding this info would require adding new columns: extending coverage to the above information sources will not require changes to the existing database schema. -->

@@

We first discuss the `childes-db` web apps and then introduce the `childesr` `R` package.

@@

The ability to easily browse and explore the CHILDES corpora is a cornerstone of the `childes-db` project.
To this end we have created powerful yet easy-to-use interactive web applications that enable users to visualize various dimensions of the CHILDES corpus: frequency counts, mean lengths of utterance, type-token ratios, and more.
All of this is doable without the requirement of understanding command-line tools.^[The LuCiD toolkit [@chang2017] provides related functionality for a number of common analyses.
In contrast to those tools, which focus on filling gaps not covered by CLAN -- e.g., the use of $n$-gram models, incremental sentence generation, and distributional word classification -- our web apps focus on covering the same common tasks as CLAN, but yielding visualizations for the web browser.]

Our web apps are built using Shiny, a software package that enables easy app construction using `R`.
Underneath the hood, each web app is making calls to our `childesr` API and subsequently plots the data using the popular `R` plotting package `ggplot2`.
A user's only task is to configure exactly what should be plotted through a series of buttons, sliders, and text boxes.
The user may specify what collection, corpus, child, age range, caregiver, etc., should be included in a given analysis.
The plot is displayed and updated in real-time, and the underlying data are also available for download alongside the plot.
All of these analyses may also be reproduced using the `childesr` package, but the web apps are intended for the casual user who seeks to easily extract developmental indices quickly and without any technical overhead.
<!-- We next walk through the three web apps provided on our website and describe their functionality in more detail.  -->

@@

Several of the most common of these measures are available in the Derived Measures app, which plots these measures across age for a given subset of data, again specified by collection, corpora, children, and speakers.
As with the Frequency Counts app, caregivers' lexical diversity measures can be plotted alongside children's.
We have currently implemented the following measures:

@@

In many cases a researcher may want to view the statistics and properties of corpora (e.g., their size, number of utterances, number of tokens) before choosing a target corpus or set of corpora for an analysis.
This web app is intended to provide a basic overview regarding the scale and temporal extent of various corpora in CHILDES, as well as give researchers insight into the aggregate characteristics of CHILDES.
For example, examining the aggregate statistics reveals that coverage in CHILDES peaks at around 30 months.
<!-- SM: note that the graphs don't currently include a grouping factor, so adding multiple corpora generates one line, so you can't actually compare two corpora in the same visualization -->

@@

Although the interactive analysis tools described above cover some of the most common use cases of CHILDES data, researchers interested in more detailed and flexible analyses will want to interface directly with the data in `childes-db`.
Making use of the `R` programming language [@R-base], we provide the `childesr` package. `R` is an open-source, extensible statistical computing environment that is rapidly growing in popularity across fields and is increasing in use in child language research [e.g. @song2015; @norrman2015].
The `childesr` package abstracts away the details of connecting to and querying  the database.
Users can take advantage of the tools developed in the popular `dplyr` package [@R-dplyr], which makes manipulating large datasets quick and easy.
We describe the commands that the package provides and then give several worked examples of analyses using the package.

The `childesr` package is easily installed via CRAN, the comprehensive `R` archive network.
To install, simply type: `install.packages("childesr")`. After installation, users have access to functions that can be used to retrieve tabular data from the database:

@@

In this section we give a number of examples of how `childes-db` can be used in both research and teaching, using both the web apps and the `R` API. Note that all of these examples use `dplyr` syntax [@R-dplyr]; several accessible introductions to this framework are available online [e.g., @wickham2016].

@@

```{r color-shiny, out.width = "500px", fig.cap = "An example of using the Frequency shiny app to explore how children's color input changes over development."}
knitr::include_graphics("images/color-shiny.png")
```

@@

For this analysis, we will define our words of interest as the basic color words of English (except for gray, which children hear very rarely).
We store these in the `colors` variable, and then use the `get_types()` function from `childesr` to get the type frequency of each of these words in all of the corpora in CHILDES.
All other functions are provided by base `R` or the `tidyverse` package. 
For demonstration, we look only at the types produced by the speakers in each corpus tagged as Mother and Father.
We also restrict ourselves to children from 1--5 years old (12--60 months), and look only at the North American English corpora.

@@

We now join these two pieces of information together -- how many times each speaker produced each color word, and how many total words they produced.
We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin. <!-- KM: The back-to-back summarise() calls feel a little opaque. I think there is an implicit group_by() step? -->
For comparability with the web app analysis, these proportions are converted to parts per million words.

@@

```{r color-plot, fig.height = 5, fig.width = 10, fig.cap ="Color frequency as a function of age. Points represent means across transcripts, error bars represent 95% confidence intervals computed by nonparametric bootstrap. Y-axes are free because non-stationarity is evaluated within each color word, while their absolute frequencies vary widely.", cache = TRUE}
ggplot(count_estimates_with_error, aes(x = age_bin, y = empirical_stat,
                            ymin = ci_lower, ymax = ci_upper,
                            color = color, label = color)) +
  geom_pointrange(position = position_dodge(.5)) +
  geom_point() +
  geom_smooth(method = "loess", se = F, span = .5) +
  facet_wrap(~ color, scales = "free_y") +
  theme_few() +
  scale_color_manual(values = graph_colors$graph_color) +
  scale_x_continuous(breaks = seq(12, 60, 12),
                     name = "Age (months)") +
  scale_y_continuous(name = "Frequency (per million words)") +
  geom_dl(method = list(dl.trans(x = x +.2), "last.qp", cex = 1)) +
  theme(legend.position = "none")
```

@@

```{r echo=TRUE}
mtld_data <- speaker_stats %>%
  filter(!is.na(target_child_sex), target_child_name != "Leo",
         language %in% included_languages) %>%
  group_by(target_child_id, target_child_age, target_child_sex, language) %>%
  summarise(measure = mean(mtld)) %>%
  ungroup() %>%
  mutate(age_years = target_child_age / 12,
         target_child_sex = factor(target_child_sex, 
                                   levels = c("male","female"))) %>%
  filter(age_years < max_age)
```

@@

```{r echo=FALSE}
mod <- lmer(measure ~ target_child_sex * poly(age_years, 2) - 1 +
                    (1 | target_child_id) +
                    (target_child_sex + age_years | language),
  data = mtld_data, REML = TRUE, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

mod_data$measure <- predict(mod, newdata = mod_data, re.form = ~(target_child_sex + age_years | language))
```

@@

## Teaching with childes-db

@@

1. introduce the research question (e.g., What are the types of words that children first produce?),
2. allow students to reflect or do a pair-and-share discussion with their neighbor,
3. show the trajectory of a single lexical item while explaining key parts of the visualization (see Panel A of Figure \ref{fig:in-class-demo-plot}),
4. elicit hypotheses from students about the kinds of words that children are likely to produce,
5. make real-time queries to the web application to add students' suggestions and talk through the updated plots (Panels B and C of Figure \ref{fig:in-class-demo-plot}), and
6. finish by entering a pre-selected set of words that communicate the important takeaway point (Panel D of Figure \ref{fig:in-class-demo-plot}) .

```{r in-class-demo-plot, fig.pos="h", out.width = "95%", fig.cap="Worked example of using the web applications for in-class teaching. Panels A-D show how an instructor could dynamically build a plot during a lecture to demonstrate a key concept in language acquisition."}

@@

We have presented `childes-db`, a database formatted mirror of the CHILDES dataset.
This database -- together with the `R` API and web apps -- facilitates the use of child language data.
For teachers, students, and casual explorers, the web apps allow browsing and demonstration.
For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data.
We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.
